---
title: "Loan Prediction Project"
author: "George Moisescu"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this project I will try to use 6 machine learnig algorithm for predicting which individuals included in the test set database will have or will not have access to credit.
I will use the data from Kaggle platform: https://www.kaggle.com/altruistdelhite04/loan-prediction-problem-dataset

```{r, echo = FALSE, message=FALSE, warning=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
if(!require(dslabs)) install.packages("dslabs", repos = "http://cran.us.r-project.org")


library(tidyverse)
library(dslabs)
library(caret)
library(scales)

loan_data <- read.csv("https://raw.githubusercontent.com/GeorgeMoisescu/ChooseYourOwn/main/train_loan.csv")

```


```{r,message=FALSE, warning=FALSE}


head(loan_data,4)
str(loan_data)

```

The dataset contain 614 observations in 13 variables.
I will drop the Loan_ID variable because it is not useful for building the prediction algorithm .

## Describing the variables

1. Gender:            the gender of applicant(male or female)
2. Married:           if the applicant is married or single
3. Dependents:        the number of dependents the applicant have(0,1,2,3+)
4. Education:         the studies of the applicant(Graduate or Not Graduate)
5. Self_Employed:     if the applicant is employed or self_employed
6. ApplicantIncome:   integer variable who tell us the applicant income
7. CoapplicantIncome: numeric variable who tell us the coapplicant income
8. LoanAmount:        the amount of the loan
9. Loan_Amount_Term:  the period in which the loan is given
10. Credit_History:   if the applicant has history credit
11. Property_Area:    the place where the property of the applicant is situated
12: Loan_Status:      status of the application.

# Exploratory Data Analysis

As we can see the type of variables are character, integer and numeric. I will transform the data into factors and integer.

```{r,message=FALSE, warning=FALSE}

loan_data$Gender <- as.factor(loan_data$Gender)
loan_data$Married <- as.factor(loan_data$Married)
loan_data$Dependents <- as.factor(loan_data$Dependents)
loan_data$Education <- as.factor(loan_data$Education)
loan_data$Self_Employed <- as.factor(loan_data$Self_Employed)
loan_data$CoapplicantIncome <- as.integer(loan_data$CoapplicantIncome)
loan_data$Credit_History <- as.factor(loan_data$Credit_History)
loan_data$Property_Area <- as.factor(loan_data$Property_Area)
loan_data$Loan_Status <- as.factor(loan_data$Loan_Status)
loan_data$Loan_Amount_Term <- as.factor(loan_data$Loan_Amount_Term)
```

Summarising the dataset:

```{r,message=FALSE, warning=FALSE}
summary(loan_data)
```

There are missing values in the Gender, Married, Dependents, Self_Employed, Credit_History variables.

## Dealing with the missing values in the loan_data

### Gender variable

```{r,message=FALSE, warning=FALSE}
gender_data <- loan_data %>% filter(Gender != "") %>% group_by(Gender) %>% 
          summarize(n = n()) %>% mutate(percentage = round(n/sum(n),digits = 2))
gender_data
```
We have 19% female applicants and 81% male applicants

```{r,message=FALSE, warning=FALSE}
missing <- loan_data %>% filter(Gender == "")
nrow(missing)
```
We have 13 applicants with no value for Gender.
19% of 13 is 2. I will fill 2 of 13 with female and 11 of 13 with male

```{r,message=FALSE, warning=FALSE}
gender_index <- which(loan_data$Gender == "")
gender_index

loan_data$Gender[c(172,461)] <- "Female"

gender_index_male <-setdiff(gender_index,c(172,461))
gender_index_male

for (i in gender_index_male) {
  loan_data$Gender[i] <- "Male"
}

loan_data$Gender <- droplevels(loan_data$Gender)
summary(loan_data$Gender)
levels(loan_data$Gender)
```
###  Married variable

```{r,message=FALSE, warning=FALSE}
married_data <- loan_data %>% filter(Married != "") %>% group_by(Married) %>%
          summarize(n = n()) %>% mutate(percentage = round(n/sum(n),digits = 2))
married_data
```
We have 65% married applicants and 35% single applicants

```{r,message=FALSE, warning=FALSE}
summary(loan_data$Married)
```
We have 3 missing values.
```{r,message=FALSE, warning=FALSE}
married_index <- which(loan_data$Married == "")
married_index

round(0.65 * length(married_index))
```
I will fill 2 observations with "Yes" and one with "No"

```{r,message=FALSE, warning=FALSE}
loan_data$Married[c(105,229)] <- "Yes"
loan_data$Married[c(436)] <- "No"

loan_data$Married <- droplevels(loan_data$Married)
summary(loan_data$Married)
```
### Dependents variable


```{r,message=FALSE, warning=FALSE}
summary(loan_data$Dependents)
dependents_data <- loan_data %>% filter(Dependents != "") %>% 
                    group_by(Dependents) %>% summarize(n = n()) %>%
                    mutate(percentage = round(n/sum(n),digits = 2))
dependents_data
```
We have 58% applicants with no dependent, 17% with 1 dependent, 17% with 2 dependents and 9% with 3 or more dependents
We have 15 applicants with no value for Dependents.

```{r,message=FALSE, warning=FALSE}
dependents_index <- which(loan_data$Dependents == "")
dependents_index

round(0.58 * length(dependents_index))
```

I will fill 9 observation with 0, 3 observations with 1 and 3 with 2.

```{r,message=FALSE, warning=FALSE}
loan_data$Dependents[c(103,227,229,302,336,356,436,572,598)] <- 0

round(0.17 * length(dependents_index))
loan_data$Dependents[c(105,121,294)] <- 1
loan_data$Dependents[c(333,347,518)] <- 2
```

```{r,message=FALSE, warning=FALSE}
loan_data$Dependents <- droplevels(loan_data$Dependents)
summary(loan_data$Dependents)
```

### Self_Employed variable

```{r,message=FALSE, warning=FALSE}
summary(loan_data$Self_Employed)
selfemp_data <- loan_data %>% filter(Self_Employed != "") %>% group_by(Self_Employed) %>% summarize(n = n(), avg_income = mean(ApplicantIncome)) %>% mutate(percentage = round(n/sum(n),digits = 2))
selfemp_data
```
We have 86% applicants who are not self_employed and 14% who are self_employed
The average income for self_employed applicant is 7381 and for those who employed is 5050
We have 32 applicants with no value for Self_Employed.


```{r,message=FALSE, warning=FALSE}
selfemp_index <- which(loan_data$Self_Employed == "")
selfemp_index

loan_data$ApplicantIncome[selfemp_index]

round(0.14 * length(selfemp_index))
```
We choose 4 applicant out of 32 with income close to average income of the category.

```{r,message=FALSE, warning=FALSE}
loan_data$Self_Employed[c(96,108,334,433)] <- "Yes"
```

The rest of 28 observation I fill with "No"

```{r,message=FALSE, warning=FALSE}
selfemp_index_no <-setdiff(selfemp_index,c(96,108,334,433))
selfemp_index_no

for (i in selfemp_index_no) {
  loan_data$Self_Employed[i] <- "No"
}

loan_data$Self_Employed <- droplevels(loan_data$Self_Employed)
summary(loan_data$Self_Employed)
```

### LoanAmount variable

```{r,message=FALSE, warning=FALSE}
loan_data %>% filter(!is.na(LoanAmount)) %>% ggplot(aes(LoanAmount)) + 
      geom_histogram(bins = 20) + ggtitle("LoanAmount distribution in dataset")
summary(loan_data$LoanAmount)
```
The distribution is right skewed and we have missing values. I will fill the NA values with median value.
```{r,message=FALSE, warning=FALSE}
median(loan_data$LoanAmount,na.rm = TRUE)
loan_data$LoanAmount[is.na(loan_data$LoanAmount)] <- 
                                      median(loan_data$LoanAmount,na.rm = TRUE)
```

### Loan_Amount_Term variable

```{r,message=FALSE, warning=FALSE}
loanamountterm_data <- loan_data %>% filter(!is.na(Loan_Amount_Term)) %>% 
                    group_by(Loan_Amount_Term) %>% summarize(n = n()) %>% 
                    mutate(percentage = round(n/sum(n),digits = 2))
loanamountterm_data
```
I am going to fill the NA's based on the proportion of each category.

```{r,message=FALSE, warning=FALSE}
loan_data %>% filter(is.na(Loan_Amount_Term)) %>%nrow()

loanamountterm_index <- which(is.na(loan_data$Loan_Amount_Term))
loanamountterm_index

round(0.85 * 14)
loan_data$Loan_Amount_Term[c(20,37,45,46,74,113,166,198,224,233,336,368)] <- 360

round(0.07 * 14)

loan_data$Loan_Amount_Term[422] <- 180
loan_data$Loan_Amount_Term[424] <- 480
```
I filled 12 of 14 observations with 360, 1 observation with 180 and 1 with 480


```{r,message=FALSE, warning=FALSE}
summary(loan_data$Loan_Amount_Term)
```
### Credit_History variable

```{r,message=FALSE, warning=FALSE}
summary(loan_data$Credit_History)
credithistory_data <- loan_data %>% filter(!is.na(Credit_History)) %>%
                      group_by(Credit_History) %>% summarize(n = n()) %>% 
                      mutate(percentage = round(n/sum(n),digits = 2))
credithistory_data
```

We have 50 observations with missing values, 16% of applicants witch don't have credit history and 84% of applicant witch have credit history.
I will fill the NA values based on proportion of each category.
```{r,message=FALSE, warning=FALSE}
credithistory_index <- which(is.na(loan_data$Credit_History))
credithistory_index

round(0.16 * 50)

credithistory_index_no <- c(17,43,96,130,220,260,318,393)
loan_data$Credit_History[credithistory_index_no] <- 0

credithistory_index_yes <- setdiff(credithistory_index,credithistory_index_no)
credithistory_index_yes

for (i in credithistory_index_yes) {
  loan_data$Credit_History[i] <- 1
}

```

```{r,message=FALSE, warning=FALSE}
loan_data <- loan_data[,-1]
```

## Dealing with outliers

I use Tukey's definition of an outlier

### Removing outliers for ApplicantIncome

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(ApplicantIncome)) + geom_histogram(bins = 20)
summary(loan_data$ApplicantIncome)
```

I will keep the values witch are between lower and higher values.

```{r,message=FALSE, warning=FALSE}
cutt_off <- 1.5 * IQR(loan_data$ApplicantIncome)
cutt_off
lower_value <- quantile(loan_data$ApplicantIncome,0.25) - cutt_off
higher_Value <- quantile(loan_data$ApplicantIncome,0.75) + cutt_off
lower_value
higher_Value

appinc_dropindex <- which(loan_data$ApplicantIncome > higher_Value)
appinc_dropindex

loan_data <- loan_data[-appinc_dropindex,]
loan_data %>% ggplot(aes(ApplicantIncome)) + geom_histogram(bins = 20)
```
### Removing outliers for CoapplicantIncome

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(CoapplicantIncome)) + geom_histogram(bins = 20)
summary(loan_data$CoapplicantIncome)
```
I will keep the values witch are between lower and higher values.

```{r,message=FALSE, warning=FALSE}
cutt_off <- 1.5 * IQR(loan_data$CoapplicantIncome)
cutt_off
lower_value <- quantile(loan_data$CoapplicantIncome,0.25) - cutt_off
higher_Value <- quantile(loan_data$CoapplicantIncome,0.75) + cutt_off
lower_value
higher_Value
coappinc_dropindex <- which(loan_data$CoapplicantIncome > higher_Value)
coappinc_dropindex
loan_data <- loan_data[-coappinc_dropindex,]
loan_data %>% ggplot(aes(CoapplicantIncome)) + geom_histogram(bins = 20)
```

## Grouping the ApplicantIncome and CoapplicantIncome

I will group the ApplicantIncome and CoapplicantIncome in a single variable TotalIncome.

```{r,message=FALSE, warning=FALSE}
loan_data <- loan_data %>% mutate(TotalIncome = ApplicantIncome + CoapplicantIncome)
loan_data <- loan_data[,-c(6,7)]
loan_data <- loan_data[,c(1:9,11,10)]
```


## Data Visualisation

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Gender, fill = Loan_Status)) + 
                          geom_bar(position=position_dodge()) + 
                          ggtitle("Gender Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Married, fill = Loan_Status)) + 
                        geom_bar(position=position_dodge()) + 
                        ggtitle("Married Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Dependents, fill = Loan_Status)) +
                      geom_bar(position=position_dodge()) + 
                      ggtitle("Dependents Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Education, fill = Loan_Status)) + 
                    geom_bar(position=position_dodge()) +
                    ggtitle("Education Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Self_Employed, fill = Loan_Status)) + 
                  geom_bar(position=position_dodge()) + 
                  ggtitle("Self Employed Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(TotalIncome,LoanAmount, color = Loan_Status)) + 
          geom_point() + ggtitle("Variation of LoanAmount with ApplicantIncome")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Loan_Amount_Term, fill = Loan_Status)) +
                geom_bar(position=position_dodge()) + 
                ggtitle("Loan Amount Term Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Credit_History, fill = Loan_Status)) + 
                geom_bar(position=position_dodge()) +
                ggtitle("Credit History Distribution with Status of the Loan")
```

```{r,message=FALSE, warning=FALSE}
loan_data %>% ggplot(aes(Property_Area, fill = Loan_Status)) +
                geom_bar(position=position_dodge()) + 
                ggtitle("Property Area Distribution with Status of the Loan")
```

## Scaling the data

I will transform the TotalIncome variable so that it will fit into the scale of 0-1

```{r,message=FALSE, warning=FALSE}
loan_data$TotalIncome <- (loan_data$TotalIncome - min(loan_data$TotalIncome)) /
                      (max(loan_data$TotalIncome) - min(loan_data$TotalIncome))
```

```{r,message=FALSE, warning=FALSE}
summary(loan_data$TotalIncome)
```

## Features encoding

I will convert categorical data into numbers because it lead us to a better model and most of the algorithm cannot handle categorical data.


```{r,message=FALSE, warning=FALSE}
loan_data$Gender <- ifelse(loan_data$Gender == "Male",1,0)
loan_data$Married <- ifelse(loan_data$Married == "Yes",1,0)


loan_data$Dependents <- as.integer(loan_data$Dependents)

indexs <- which(loan_data$Dependents == 1)
loan_data$Dependents[indexs] <- 0
indexs <- which(loan_data$Dependents == 2)
loan_data$Dependents[indexs] <- 1
indexs <- which(loan_data$Dependents == 3)
loan_data$Dependents[indexs] <- 2
indexs <- which(loan_data$Dependents == 4)
loan_data$Dependents[indexs] <- 3

loan_data$Education <- ifelse(loan_data$Education == "Graduate",1,0)
loan_data$Self_Employed <- ifelse(loan_data$Self_Employed == "Yes",1,0)
loan_data$LoanAmount <- (loan_data$LoanAmount - min(loan_data$LoanAmount)) / 
                      (max(loan_data$LoanAmount) - min(loan_data$LoanAmount))
loan_data$Loan_Amount_Term<-as.numeric(as.character(loan_data$Loan_Amount_Term))
loan_data$Loan_Amount_Term <- (loan_data$Loan_Amount_Term - 
          min(loan_data$Loan_Amount_Term)) / (max(loan_data$Loan_Amount_Term) - 
          min(loan_data$Loan_Amount_Term))
loan_data$Credit_History <- as.numeric(as.character(loan_data$Credit_History))

levels(loan_data$Property_Area)
loan_data$Property_Area <- as.character(loan_data$Property_Area)
indexs <- which(loan_data$Property_Area == "Rural")
loan_data$Property_Area[indexs] <- 0
indexs <- which(loan_data$Property_Area == "Semiurban")
loan_data$Property_Area[indexs] <- 1
indexs <- which(loan_data$Property_Area == "Urban")
loan_data$Property_Area[indexs] <- 2
loan_data$Property_Area <- as.numeric(loan_data$Property_Area)
```

```{r,message=FALSE, warning=FALSE}

head(loan_data,7)

```

# Modeling approach

For this project I will use 6 machine learning algorithms:

* k-nearest neighbors
* Generalized Linear Model
* Quadratic Discriminant Analysis
* Linear discriminant analysis
* Decision Trees
* Random forest

Splitting the loan_data in train data and test data:

```{r,message=FALSE, warning=FALSE}

set.seed(1)
index <- createDataPartition(loan_data$Loan_Status,times = 1,p = 0.8, 
                             list = FALSE)
loan_data_train <- loan_data[index,]
loan_data_test <-loan_data[-index,]

```

## K-nearest neighbors

```{r,message=FALSE, warning=FALSE}

control <- trainControl(method = "cv", number = 10, p = .9)
train_knn <- train(Loan_Status ~ ., method = "knn", data = loan_data_train, 
                   tuneGrid = data.frame(k = seq(9, 71, 2)),trControl = control)
ggplot(train_knn, highlight = TRUE)
train_knn$bestTune
pred_LoanStatus_knn <- predict(train_knn,loan_data_test)

```

Accuracy of the model:

```{r,message=FALSE, warning=FALSE}

knn_accuracy <- confusionMatrix(pred_LoanStatus_knn,loan_data_test$Loan_Status)$
                                                          overall["Accuracy"]
knn_accuracy
```

## Generalized Linear Model

```{r,message=FALSE, warning=FALSE}

train_glm <- train(Loan_Status ~ ., method = "glm", data = loan_data_train)
pred_LoanStatus_glm <- predict(train_glm,loan_data_test)

```

Accuracy of the model:

```{r,message=FALSE, warning=FALSE}

glm_accuracy <- confusionMatrix(pred_LoanStatus_glm,loan_data_test$Loan_Status)$
                                                          overall["Accuracy"]
glm_accuracy
```

## Quadratic Discriminant Analysis

```{r,message=FALSE, warning=FALSE}
train_qda <- train(Loan_Status ~ ., method = "qda", data = loan_data_train)
pred_LoanStatus_qda <- predict(train_qda,loan_data_test)
```

Accuracy of the model:

```{r,message=FALSE, warning=FALSE}
qda_accuracy <- confusionMatrix(pred_LoanStatus_qda,loan_data_test$Loan_Status)$
                                                          overall["Accuracy"]
qda_accuracy
```

## Linear discriminant analysis

```{r,message=FALSE, warning=FALSE}
train_lda <- train(Loan_Status ~ ., method = "lda", data = loan_data_train)
pred_LoanStatus_lda <- predict(train_lda,loan_data_test)
```

Accuracy of the model:

```{r,message=FALSE, warning=FALSE}
lda_accuracy <- confusionMatrix(pred_LoanStatus_lda,loan_data_test$Loan_Status)$
                                                      overall["Accuracy"]
lda_accuracy
```

## Decision Trees

```{r,message=FALSE, warning=FALSE}
train_rpart <- train(Loan_Status ~ ., method = "rpart",data = loan_data_train)
pred_LoanStatus_rpart <- predict(train_rpart,loan_data_test)
```

Accuracy of the model:

```{r,message=FALSE, warning=FALSE}
rpart_accuracy <- confusionMatrix(pred_LoanStatus_rpart,loan_data_test$
                                          Loan_Status)$overall["Accuracy"]
rpart_accuracy

```

## Random forest

```{r,message=FALSE, warning=FALSE}
train_rf <- train(Loan_Status ~ ., method = "rf", data = loan_data_train)
pred_LoanStatus_rf <- predict(train_rf,loan_data_test)
```

Accuracy of the model:

```{r,message=FALSE, warning=FALSE}
rf_accuracy <- confusionMatrix(pred_LoanStatus_rf,loan_data_test$Loan_Status)$
                                                        overall["Accuracy"]
rf_accuracy
```
# Models performance

```{r,message=FALSE, warning=FALSE}
model <- rep(0,6)
accuracy <- rep(0,6)
accuracy_summary <- data.frame(model, accuracy)
accuracy_summary$model[1] <- "knn"
accuracy_summary$accuracy[1] <- round(knn_accuracy,4)
accuracy_summary$model[2] <- "glm"
accuracy_summary$accuracy[2] <- round(glm_accuracy,4)
accuracy_summary$model[3] <- "qda"
accuracy_summary$accuracy[3] <- round(qda_accuracy,4)
accuracy_summary$model[4] <- "lda"
accuracy_summary$accuracy[4] <- round(lda_accuracy,4)
accuracy_summary$model[5] <- "rpart"
accuracy_summary$accuracy[5] <- round(rpart_accuracy,4)
accuracy_summary$model[6] <- "rf"
accuracy_summary$accuracy[6] <- round(rf_accuracy,4)

accuracy_summary %>% ggplot(aes(x = model,y = accuracy)) +
  geom_bar(stat = "identity") + scale_y_continuous(limits=c(0.7,0.9),oob =
              rescale_none) + geom_text(aes(label = accuracy), vjust = -1)

```

# Conclusions

We can see that the most accurate models for our dataset is GLM, LDA and LPART with an accuracy of 0.8349.

